%% -*- mode: LaTeX; compile-command: "mk" -*-
\documentclass{article}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Packages

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage[all]{xy}
\usepackage{prettyref}
\usepackage{mdframed}
\usepackage{todonotes}
\usepackage{xspace}

\usepackage[backend=pgf,extension=pgf,input,outputdir=diagrams]{diagrams-latex}

\usepackage[authoryear]{natbib}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Theorem-like environments

\newtheorem{thm}{Theorem}[section]
\newtheorem{prop}[thm]{Proposition}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{cor}[thm]{Corollary}
\newtheorem{conj}[thm]{Conjecture}

\theoremstyle{definition}

\newtheorem{defn}[thm]{Definition}
\newtheorem{ex}{Example}

\theoremstyle{remark}
\newtheorem*{rem}{Remark}
\newtheorem*{nota}{Notation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Commentary environment

\newmdenv[leftmargin=1em, rightmargin=1em]{commentary}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Math formatting

\newcommand{\B}{\mathbb{B}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\N}{\mathbb{N}}
\DeclareMathOperator{\el}{el}

\newcommand{\bij}{\stackrel{\sim}{\rightarrow}}
\newcommand{\inj}{\hookrightarrow}

\DeclareMathOperator{\Card}{Card}
\DeclareMathOperator{\Aut}{Aut}
\newcommand{\Mon}{\mathrm{Mon}}
\newcommand{\unl}[1]{\tilde{#1}}

\newcommand{\term}[1]{\emph{#1}}

\newcommand{\union}{\cup}
\newcommand{\intersect}{\cap}

\newcommand{\pointed}[1]{#1^{\bullet}}
\newcommand{\Gc}{\pointed{G_c}}
\newcommand{\Gcc}{\pointed{G_{cc}}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Prettyref

\newcommand{\Sect}{\S}

\newrefformat{fig}{Figure~\ref{#1}}
\newrefformat{chap}{Chapter~\ref{#1}}
\newrefformat{sec}{\Sect\ref{#1}}
\newrefformat{eq}{equation~\eqref{#1}}
\newrefformat{prob}{Problem~\ref{#1}}
\newrefformat{tab}{Table~\ref{#1}}
\newrefformat{thm}{Theorem~\ref{#1}}
\newrefformat{lem}{Lemma~\ref{#1}}
\newrefformat{prop}{Proposition~\ref{#1}}
\newrefformat{defn}{Definition~\ref{#1}}
\newrefformat{cor}{Corollary~\ref{#1}}
\newrefformat{conj}{Conjecture~\ref{#1}}
\newrefformat{ex}{Example~\ref{#1}}

\newcommand{\pref}[1]{\prettyref{#1}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Other formatting

\newcommand{\etc}{\textit{etc.}\xspace}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\section{Introduction}
\label{sec:introduction}

\todo{Finish translating introduction}

\section{Species of Structures}
\label{sec:species-of-structures}

There already exists a precise concept of species of structures (\todo{Bourbaki
[2]}). Describing a particular species is often done by specifying the
conditions which a structure must satisfy to belong to the species.
This description may take the form of an axiomatic theory of the species
being considered. A key part of the concept is the transport of structures.

We will abstract the concept of species so that the transport of
structures is the main aspect. Moreover, as we only deal
with the problems of counting and finite enumeration, we will confine ourselves
to finitary species, unless otherwise noted.

\begin{commentary}
  See \citet[pp. 4 XXX]{BLL} for some examples of ``axiomatic theories
  of species''.  XXX more examples.  The main point here is about
  \emph{transport} of species, that is, the ability to ``swap out''
  the labels in a structure for some other labels.  In most previous
  treatments this was taken as something secondary, a desideratum to
  be verified after each definition; Joyal's insight is to make
  transport itself the central defining feature of species.
\end{commentary}

\subsection{Species and cardinality}
\label{sec:species-and-cardinality}

\begin{defn}
  A (finitary) species is an endofunctor $M: \B \to \B$
  on the groupoid $\B$ of finite sets and bijections.
\end{defn}

If $E$ is a finite set, $M [E]$ is the set of all \term{structures} of
the species $M$ on $E$. We say that $E$ is the \term{underlying} set
of $s \in M[E]$, or also that it is \term{supported} by $E$. We say
also, in an abuse of language, that $s$ is an \term{element} of $M$
($s \in M$) and that it is an \term{$M$-structure}.

\begin{commentary}
  Commentary explaining $\B$.  Remind what a groupoid is.  Can
  probably steal some stuff from my dissertation.  Show what $\B$
  looks like.  Note that any groupoid in fact looks like a disjoint
  union of cliques.

  The fact that a species is a \emph{functor} corresponds to the idea
  of ``baking in'' the notion of \emph{transport}: functors send not
  only objects to objects (in this case, sets of labels to sets of
  structures) but also morphisms to morphisms (in this case,
  bijections between label sets to bijections between labelled
  structures).

  Many modern treatments define species as functors $\B \to \E$, where
  $\E$ is the category of (finite) sets and total functions (XXX cite
  some thing(s)).  Note, however, that functors necessarily preserve
  isomorphisms, so every bijection in $\B$ must actually map to a
  bijection in $\E$.  Thus, for most purposes the distinction is not
  an important one.  In fact, functors $\B \to \E$ are introduced
  later, in \pref{sec:category-of-species}.
\end{commentary}

If $u: E \to F$ is a bijection, the element $t = M [u] (s)$ is the
structure on $F$ obtained by \term{transport along $u$}. The bijection
$u$ is an \term{isomorphism} between $s$ and $t$:
\[ u : s \to t. \]
\begin{commentary}
  Note that $u : s \to t$ is a mild abuse of notation; $u$ is, of
  course, a bijection between label sets $E$ and $F$.  The point is
  that, via transport, $u$ induces a bijection between the sets of
  structures $M[E]$ and $M[F]$.  If $s \in M[E]$ and $t \in M[F]$ are
  related by this bijection then we say that $s$ and $t$ are
  isomorphic.

  A concrete example should help to make the idea clear.  XXX example
  here of isomorphic trees, as well as non-isomorphic trees.
\end{commentary}
We denote by $\el (M)$ the category whose objects are the $M$-structures and
whose morphisms are isomorphisms of $M$-structures; it is the
groupoid of \term{elements} of $M$.

There is a \term{forgetful}
functor $U: \el (M) \to \B$ whose
value on $s \in M$ is the \term{underlying} set of the structure $s$. The concept
of isomorphism of structures defines an equivalence relation whose
classes are the \term{types} of structures of the species M; these classes are
the \term{connected components} of the groupoid $\el (M)$. We use the notation
$\pi_0 (M)$ to denote the set of types (of structures) of the species $M$. If
$s \in M$, we denote the \term{type} of $s$ by the notation $|s| \in \pi_0 (M)$.

\newcommand{\Simp}{\mathcal{S}}

\begin{ex}
  Recall that a \emph{simplicial scheme} structure on $E$ is a set
  $\Simp$ of non-empty subsets of $E$ such that (i) every non-empty
  subset contained in an element of $\Simp$ belongs to $\Simp$, (ii)
  the singletons $\{x\}$ for $x \in E$ belong to $\Simp$. The elements
  of $\Simp$ are \emph{simplices}. The dimension of a simplex is one
  less than its cardinality. A \emph{graph} is a simplicial scheme
  whose simplices have dimension $\leq 1$. If $u : E \to F$ is a
  bijection, it is clear that $u (\Simp) = \{u(S) \mid S \in \Simp\}$
  is also a simplicial scheme structure on $F$. We can therefore
  consider the \emph{species} of simplicial schemes. It is also clear
  that if $\Simp$ is a graph then $u(\Simp)$ is one too; we obtain the
  species of graphs, which is a \emph{subspecies} of the species of
  simplicial schemes. More generally, any property $P$ which applies
  to simplicial schemes, and whose satisfaction \todo{wording?} is
  invariant under isomorphism, determines a subspecies of the species
  of simplicial schemes. For example, connectedness is such an
  invariant property. The species of \emph{forests} is that of graphs
  \emph{without cycles}, the species of \emph{trees} is that of
  connected forests, etc.
\end{ex}

\begin{ex} \label{ex:endofunctions}
  The transport of an endofunction $a : E \to E$ along a bijection $u
  : E \bij F$ is by \emph{conjugation}: $\phi \mapsto u \phi
  u^{-1}$. The species of \emph{permutations} is a subspecies of the
  species of endofunctions. If we require that the graph of an
  endofunction is connected, we obtain the subspecies of
  \emph{connected endofunctions} and, likewise, that of \emph{circular
    permutations}. An important concept is that of \emph{contraction}:
  an endofunction $\phi : E \to E$ is a contraction if there exists
  $x_0 \in E$ such that for every $x \in E$ we have $\phi^n(x) = x_0$
  when $n$ is large enough. That is, a contraction is an endofunction
  which is ultimately constant.
\end{ex}

\begin{ex}
  Let $S$ be the species of permutations. Consider the groupoid $\el
  (S)$ of elements of $S$. The objects of $\el (S)$ are the sets $E
  \in \B$ equipped with a permutation $\sigma_E \in S [E]$. The
  morphisms $(E, \sigma_E) \to (F, \sigma_F)$ are the bijections $u: E
  \to F$ such that $u\sigma_E = \sigma_F u$. Let $x_1, x_2, x_3,
  \dots$ be an infinite sequence of variables.  Let $I (\sigma_E) =
  x_1^{d_1} \dots x_n^{d_n}$, where $n = \Card E$ and where $d_i$ is
  the number of cycles of length $i$ in $\sigma_E$. Two objects $(E,
  \sigma_E)$ and $(F, \sigma_E)$ in $\el (S)$ are isomorphic if and
  only if $I (\sigma_E) = I (\sigma_F)$. The set $\pi_0(S)$ of
  connected components of the groupoid $\el (S)$ is therefore
  naturally identified with the set $\Mon(x)$ of all monomials in the
  variables $x_1, x_2, x_3, \dots$
\end{ex}

\subsubsection{}
The group $E!$ of permutations of $E$ acts on $M [E]$ by transport
of structures. The set $\pi_0 (M [E])$ of \emph{orbits} is identified
with the set of types of $M$-structures supported by sets equipotent
with $E$. We identify the orbit of $s \in M[E]$ with its type
$|s|$. The \emph{stabilizer} subgroup of an element $s \in M [E]$ is
the group $\Aut (s)$ of \emph{automorphisms} of $s$. We have the
well-known formula \[ \Card |s| = \frac{n!}{\Card \Aut(s)}. \]

One of the fundamental problems of enumerative combinatorics is to
evaluate the two infinite sequences of numbers
\begin{gather}
  \Card M[n], \quad \text{$n \geq 0$ ($[n] = \{1, 2, \dots, n\}$)}, \\
  \Card \pi_0(M[n]), \quad n \geq 0.
\end{gather}

We define two generating functions. The first is a series of Hurwitz
(Comtet [6] \todo{citation}):

\begin{equation}
  M(x) = \sum_{n \geq 0} \Card M[n] \frac{x^n}{n!}.
\end{equation}

The second is a power series with integer coefficients (without
factorial):

\begin{equation}
  \unl M(x) = \sum_{n \geq 0} \Card \pi_0(M[n]) x^n.
\end{equation}

We say that $M (x)$ is the \emph{cardinality} of $M$. Let us see
immediately that the calculation of $\unl M (x)$ boils down to
computing the cardinality of the \emph{associated} species $\unl M$.

\begin{defn}
  A structure of the species $\unl M$ is a pair $(\sigma, s)$ where
  $\sigma$ is an automorphism of $s \in M$.
\end{defn}

\begin{prop}
  We have
  \[ \unl M (x) = \Card {\unl M}. \]
\end{prop}

\begin{proof}
  We use Burnside's lemma: if a finite group $G$ acts on a finite set
  $X$, then the cardinality of the set $\pi_0(X)$ of orbits of $X$ is
  equal to that of the set $\{(\sigma, x) \mid \sigma \in G, x \in X,
  \sigma \cdot x = x\}$ \emph{divided by} $\Card G$. (See Burnside
  [3]. \todo{Citation})
\end{proof}

\subsection{The category of species}
\label{sec:category-of-species}

Species form a category: they are functors, and one can take natural
transformations as morphisms. As it is desirable to have a larger
class of morphisms than that of isomorphisms, it is best to consider a
species as a functor $M: \B \to \E$ to the category $\E$ of finite
sets and \emph{functions} (by composing with the inclusion $\B \inj
\E$).

\begin{defn}
  A \emph{morphism} $\alpha : M \to N$ is a natural transformation
  from $M$ to $N$, considered as functors from $\B$ to $\E$.
\end{defn}

One can interpret $\alpha$ as follows: one has a \emph{construction}
$\alpha$ allowing one to produce a structure of the species $N$
(output) from a structure of the species $M$ (input), and for every
bijection $u : E \to F$ the rectangle \[ \xymatrix{M[E]
  \ar[r]^{\alpha_E} \ar[d]_{M[u]} & N[E] \ar[d]^{N[u]} \\ M[F]
  \ar[r]^{\alpha_F} & N[F] } \] commutes; this means that the
construction is \emph{equivariant} (or invariant): it does not change
if one \emph{simultaneously} transports the input and the output along
the same bijection; the vast majority of mathematical constructions
have this property.

If $\sigma_E$ is invertible regardless of $E$, the morphism $\alpha$
is an \emph{isomorphism} between $M$ and $N$. In this case, we write
$M \stackrel{\alpha}{=} N$, or more simply (in an abuse of notation) $M
= N$. If $M$ and $N$ satisfy the weaker condition $\Card M = \Card N$,
we say that $M$ and $N$ are \emph{equipotent} species, and we write $M
\equiv N$.

\begin{ex}
  The construction of the transitive closure of a graph determines a
  morphism from the species of graphs to the species of partitions.
\end{ex}

\begin{ex}
  A \emph{rooted tree} is a tree equipped with a \emph{root} (which is
  an arbitrary vertex of the underlying set). We usually orient the
  edges of a rooted tree in the direction of the root. If we adjoin a
  loop to the root, we obtain the graph of a \emph{contraction}
  (\pref{ex:endofunctions}). There is an \emph{isomorphism} between
  the species of rooted trees and the species of contractions.
\end{ex}

\begin{ex}
  The species of linear orders, of permutations, of permutations
  equipped with a fixed point, and of circular permutations equipped
  with an automorphism are all \emph{equipotent}, without being
  isomorphic.
\end{ex}

\subsubsection{} A morphism of species $M \to N$ determines a functor
$\el (M) \to \el (N)$ between the corresponding groupoids. Note that
this functor commutes with the forgetful functors \[ \xymatrix{\el(M)
  \ar[rr] \ar[dr]_U & & \el(N) \ar[dl]^U \\ & B} \] \todo{Make this
  not as wide}

It is not true that a functor $\el (M) \to \el (N)$ is always induced
by a morphism of species $M \to N$. For example, if $M$ is the species
of preorders and $N$ the species of the orders, the usual construction
of an order relation to from a pre-order (on a quotient of the
pre-order's underlying set) determines a functor $\el (M) \to \el (N)$
that \emph{does not come} from a morphism of species $M \to
N$. However, it is easily checked that every functor $\el (M) \to \el
(N)$ which commutes with the forgetful functors $U$ is induced by one
and only one species morphism $M \to N$.

\subsection{Relative species}

We want to examine the concept of a \emph{relative} species. We begin
with an example. Let $G$ be the species of \emph{graphs}. The concept
of \emph{orientation} gives us a functor $O : \el (G) \to \E$, because
one can transport a graph orientation along a graph isomorphism. The
species of \emph{orientations} (of a graph) is \emph{relative} to that
of graphs. On the other hand, the species $GO$ of \emph{oriented
  graphs} is equipped with a projection $GO \to G$.

\begin{defn}
  Let $M$ be a species. A species \emph{relative} to $M$ is a functor
  $T_M : \el (M) \to \E$.
\end{defn}

\todo{Ugh, this paragraph will need a lot of commentary and/or some
  nice pictures.}
Given $T_M$, one can construct a species $T$
equipped with a morphism $T \to^p M$: set $T [E] = \{(s, \alpha) \mid
s \in M [E], \alpha \in T_M[s]\}$. To transport $(s, \alpha) \in T[E]$
along a bijection $u : E \to F$ we begin by transporting $s$ to obtain
$t = M [u] (s)$, which gives first an isomorphism $s \to^u t \in
\el(M)$ and then $\beta = T_M [u] (\alpha)$; we set $T [u] (s, \alpha)
= (t, \beta)$. The morphism $p : T \to M$ is the projection $p (s,
\alpha) = s$. Conversely, given a morphism $T \to^p M$, we can
construct $T_M$: if $s \in M [E]$, we have $p_E: T [E] \to M [E]$ and
set $T_M [s] = p_E^{-1}\{s\} \subseteq T [E]$. Naturality of $p$
allows us to verify that if $u : E \to F$ is an isomorphism between $s
\in M [E]$ and $T \in M [F]$ then $T [u]$ turns $p_E^{-1} \{s\}$ into
$p_E^{-1} \{t\}$, which gives $T_M [u] : T_M [s] \to T_M [t]$. We
have, in fact, a precise proposition: a \term{species over $M$} is a
species $T$ equipped with a morphism $T \to^p M$. A morphism $(T, p)
\to (T', p')$ between species above $M$ is an arrow $T \to^u T'$ such
that $p'\ u = p$.

\begin{prop}
The constructions described above define a
equivalence between the category of species relative to $M$ and the category
$E \|X\|/_M$ of species over $M$.See REFERENCE 2.0 for the notation
$E\|X\|$.)
\end{prop}

Suppose $T_M: \el (M) \to E$ is given. We often say that $(s, \alpha)
\in T [E]$ is an $M$-structure $s$ \term{equipped} with an element
$\alpha \in T_M[s]$. For example, a directed graph is a graph
\term{equipped} with an orientation. A structure of the species
$\tilde M$ (DEFINITION 2) is an $M$-structure \term{equipped} with an
automorphism.  \emph{etc.}

We sometimes use the term ``enriched'' rather than ``equipped''. Thus,
if $R$ is any species, we will say that endofunction $\phi : E \to E$
is \term{$R$-enriched} if each of its \term{fibers} $\phi^{-1}\{x\}$,
$x \in E$ is equipped with an $R$-structure.  Similarly, let $a$ be an
rooted tree on $E$. The \term{fiber} $a^{-1}\{x\}$ of a
vertex $x \in E$ is the set of vertices of $a$ connected to $x$ by an
edge adjacent to $x$ (for the orientation of a tree as described in
Example XXX \todo{fill in}). We say that $a$ is \term{$R$-enriched} if each of its
fibers is equipped with an $R$-structure. (Keeping in mind the empty
fibers.)

In graphical representations of endofunctions or $R$-enriched trees it
is often convenient to assume that $R$-structures on the fibers are
placed on the set of \emph{edges} of the fibers.  For example, an
$R$-enriched tree can be represented as in \pref{fig:enriched-tree}
where an arc cutting through the edges of a fiber denotes an
$R$-structure. Don't forget the leaves.
\begin{figure}
  \centering
  \begin{diagram}[width=300]
import Diagrams
dia = drawEnrichedTree (layoutEnrichedTree figure0) # frame 0.5
  \end{diagram}
  \caption{An $R$-enriched tree} \label{fig:enriched-tree}
\end{figure}

\section{The Combinational Operations}
\label{sec:combinatorial-operations}

\newcommand{\Poly}[2]{#1 \llbracket #2 \rrbracket}

The category of species is rich in various operations. In this
section, we describe several operations of which three are binary.
The first two are the sum (disjoint) and the product. With these two
operations, the category of species becomes a kind of semi-ring.  More
precisely, let $R$ be a commutative ring, and denote by $\Poly R x$
the ring of Hurwitz series with coefficients in $R$: these are the
formal series
\[ \sum_{n \geq 0} a_n \frac{x^n}{n!}, \quad \text{where $n \geq 0,
  a_n \in R$.} \] The continued analogy here is that the category of
species would be the semiring $\Poly{\E}{X}$ of Hurwitz series, but
\emph{with coefficients from the category $\E$ of finite sets}. The
concept of cardinality induces a \emph{homomorphism}
\[ \Card : \Poly{\E}{X} \to \Poly{\Z}{x}. \]

In addition, the evaluation $M \mapsto M [0]$ is a functor preserving sum and
product
\[ \Poly \E X \to \E \]
whose kernel $J$ is an ideal on which we will describe the operations
of \term{divided powers} (Cartan [4] XXX look this up)
\[ \gamma_n : J \to \Poly \E X \quad \text{($n \geq 0$)} \]
so that we have
\[ \Card \gamma_n(M) = \frac{M(x)^n}{n!} \quad \text{($n \geq 0$)}. \]

Using these operations of divided powers we then describe
the operation of \term{substitution} of one species into another. We end this
chapter with an introduction to the \term{differential calculus} of species
and a \emph{combinatorial} proof of the Lagrange inversion formula.

\subsection{Sum and product}

The disjoint sum of two species $M$ and $N$ is the \emph{coproduct}
in the category of species:
\[ (M + N) [E] = M [E] + N [E]. \]

More generally, an arbitrary family of species $(M_i)_{i \in I}$ is
summable if for any finite set $E$, the set of indices $i \in I$ for
which $M_i[E] \neq \varnothing$ is finite.  We set
\[ \left( \sum_{i \in I} M_i \right) [ E ] = \sum_{i \in I} M_i[E]. \]
It is clear that $\Card$ preserves sum. We now turn to the definition
of the \term{product} $M \cdot N$ of two species $M$ and $N$. Define
first a \term{partition} of a set $E$ into two \term{parts} is a pair
$(E_1, E_2)$ such that $E_1 \union E_2 = E$ and $E_1 \intersect E_2 =
\varnothing$. One defines in the same way the concept of a partition
of $E$ into $n$ pieces ($n \in \N$): we write $E = E_ + \dots + E_n$
to indicate that $(E_1, \dots, E_n)$ is a partition of $D$ into $n$
pieces.

\begin{defn}
  A structure of the species $M \cdot N$ on $E \in \B$ is a quadruplet
  $(E_1, E_2, s, t)$, where $E = E_1 + E_2$ and $(s, t) \in M[E_1]
  \times N [E_2]$.
\end{defn}

\begin{prop}
We have
\[ \Card (M \cdot N) = \Card (M) \cdot \Card (N). \]
\end{prop}
\begin{proof}
By definition,
\[ (M \cdot N)[E] = \sum_{E_1 + E_2 = E} M[E_1] \times N[E_2]. \]
If $\Card E = n$ and $0 \leq k \leq n$ there are $\binom n k$
partitions $E = E_1 + E_2$ with $\Card E_1 = k$.  As a result,
\[ \Card(M \cdot N) [n] = \sum_{k=0}^n \binom n k \Card M [k] \Card N
[n - k]. \]
\end{proof}

The \term{uniform} species is a species with only one structure on each
set; one can give it various representations: the structures of
complete graphs, chaotic topologies XXX ?, and identical applications XXX? determine
the uniform species.

\begin{ex}
Let $S$ be the species of \term{permutations}, and $S_0$ that of
permutations without \emph{fixed points}, and $U$ the uniform species. We have $S = S_0 \cdot U$
XXX Fig.; taking cardinalities, we get:
\[ \frac{1}{1-x} = S_0 (x) e^x \]
and therefore
\[ S_0 (x) = \frac{e^{-x}}{1 - x}. \]
\end{ex}

\begin{ex}
  Let $D$ be the species of endofunctions, $D_0$ the species
  endofunctions equipped with a fixed point, and $A$ and the species
  of rooted trees?. Then
\[ D_0 = A \cdot D. \]
Indeed, one can partition the domain $E$ of an endofunction $\phi$ equipped with a
fixed point $x_0$ into two parts $E = E_1 + E_2$. The first, $E_1$, consists of
all the points ultimately transformed into $x_0$ by $\phi$. On $E_1$, $\phi$ induces
a contraction (XXX Chap. 1 Ex. 2), which is equivalent to a rooted tree. On
the second part $E_2$, $\phi$ induces an arbitrary endofunction (XXX
Fig. 2).
\end{ex}

The \emph{product} $M = \prod_{i=1}^n M_i$ of a finite sequence of
species can be explicitly defined as follows: a structure of the
species $M$ on $E$ is a \emph{partition} $E = E_1 + \dots + E_n$ where
each part $E_i$ (possibly empty) is \emph{equipped} with a structure
of the species $M_i$.

FIGURE 2 XXX

One can also describe the \term{power} $N^S$ of a species $N$ by a
finite set $S$: a structure of the species $N^S$ on $E$ is a function
$\chi : E \to S$ where each fiber is \emph{equipped} with a structure
of the species $N$; in other words, it is an \emph{$N$-enriched}
function.


\begin{ex}[Joyal]
  A \emph{vertebrate} is a tree \term{bipointed} by a pair $(p_0,
  p_1)$ of vertices. We say that $p_0$ is the \term{tail vertex} and
  $p_1$ the \term{head vertex}. The shortest path from the tail vertex
  to the head vertex is the \term{spine}. The vertices along the spine
  are \term{vertebrae}. For each vertex $p$, let $v(p)$ be the
  vertebra closest to $p$. The function $v$ is idempotent, and there a
  tree???-structure on each fiber of $v$.  The roots of these trees
  are the vertebrae. Thus, a vertebrate on $E$ determines a variable
  length partition $E = E_1 + \dots + E_n$, in which each part $E_i$
  is equipped with a tree structure, and vice versa. So we have the
  identity:
  \[ V = A + A^2 + A^3 + \dots, \] where $V$ is the species of
  vertebrates and $A$ that of trees??? (Fig 3 XXX).
\end{ex}

\begin{ex}
  Let $L$ be the species of linear orders and $S$ a finite set.  A
  structure of the species $L^S$ on $E$ is a function $E \to S$ where
  each fiber is equipped with a linear order. The number $l(n, s)$ of
  such objects (if $n = \Card E$ and $s = \Card S$) is the coefficient
  of $x^n/n!$ in the series $(1-x)^{-s}$. This shows that
  \[ l(n, s) = s (s + 1) \dots (s + n-1). \]
\end{ex}

FIGURE 3

Hereafter, we will often consider that a finite set $A$
determines a species by setting
\[ A [E] = \begin{cases} A \quad \text{if $E = \varnothing$} \\ \varnothing \quad \text{otherwise}. \end{cases} \]
With this convention, the category $\E$ acts as a ring of coefficients
for $\Poly \E X$, because the disjoint sum and Cartesian product of sets
can be conflated with the sum and product of species as described
earlier.

\subsection{Divided powers and substitution}

\begin{defn}
  Let $N$ be a species such that $N [0] = \varnothing$ and let $E$ be
  a finite set. An \term{assembly} of structures of the species $N$ on
  $E$ is a partition of $E$ where each class is equipped with a
  structure of the species $N$.  A \term{member} of the assembly is a
  class equipped with the corresponding $N$-structure. The
  \term{divided power} $\gamma_n(N)$ is the species of assemblies of
  $N$-structures with exactly $n$ members. The \term{exponential}
  $\exp (N)$ is the species of all assemblies of $N$-structures.
\end{defn}

\begin{prop}
We have
\begin{align*}
\Card \gamma_n(N) &= \frac{N(x)^n}{n!}, \\
\Card \exp (N) &= \exp (N (x)).
\end{align*}
\end{prop}

\begin{proof}
It obviously suffices to prove the first identity.  Note first
that a structure of the species $N^n$ on $E$ determines a partition $E
= E_1 + \dots + E_n$ into \emph{non-empty} (and disjoint) parts:
indeed, $E_i$ is equipped with an $N$-structure and by hypothesis $N
[0] = \varnothing$. If we forget about the linear order on the parts,
we have a partition $\{E_1, \dots, E_n\}$ where each class is equipped
with an $N$-structure. We have shown that an $N^n$-structure is none
other than an $N$-assembly whose members have been placed in a
\emph{linear order}:
\[ \Card N^n = n! \Card \gamma_n(N). \]
\end{proof}

For many species, one may indicate a concept of \term{connectedness}
and demonstrate that any structure consists of a partition
where each class is equipped with a connected structure. Under these conditions, a
species $M$ is isomorphic to the species of assemblies of connected structures:
$M = \exp (M_C)$.

\begin{ex}
  \term{Forests} are assemblies of trees. Forests of rooted trees are
  assemblies of rooted trees. Permutations are assemblies of circular
  permutations. Partitions are assemblies of partitions with a single
  class, \etc
\end{ex}

The operation of \term{substitution} of one species into another is
the richest in possibilities.

\begin{defn}
Let $R$ and $N$ be species and assume that $N [0] = \varnothing$.
The species $R (N)$ is that of pairs $(a, \rho)$, where $a$ is an assembly of $N$-
structures and $\rho$ is an $R$-structure on the set of members of $a$.

We say that $R (N)$ is the result of \term{substituting} $N$ into $R$. We
say that an element of $R (N)$ is an $R$-assembly (of $N$-structures). Note
immediately that $\exp (N)$ is the result of substituting $N$ in
the \emph{uniform} species (Ex. 7).
\end{defn}

\begin{thm}
  Assuming $N [0] = \varnothing$, we have
  \[ \Card R (N) = R (N (x)). \]
\end{thm}

\begin{proof}
For each integer $n \geq 0$, let $R_n$ be the species of $R$-structures
where the underlying set has cardinality $n$. It has a decomposition
as a disjoint sum
\[ R = \sum_{n \geq 0} R_n, \]
inducing a decomposition of $R$-assemblies according to the number of
members:
\[ R (N) = \sum_{n \geq 0} R_n(N). \]
An element of $R_n(N)$ is an assembly of $n$ members \emph{equipped} with an $R$-structure.
We have therefore
\[ \Card R_n(N) = \Card \gamma_n(N) \times \Card R [n], \]
resulting in
\begin{align*} \Card R(N) &= \sum_n \geq 0 \Card R[n]
  \frac{N(x)^n}{n!} \\
  &= R(N(x)).
\end{align*}
\end{proof}

\begin{rem}
  To think combinatorially it is necessary to give visual
  representations. To grasp the nature of a species is to be capable
  of representing the general shape of its structures. The
  \emph{shape} of a structure is invariant under isomorphism. What is
  needed to get represent first is the general type of structures of a
  given species. XXX ???  This type is independent of a labelling or
  an enumeration of the vertices of the underlying set. For example,
  suppose that we want to represent the general type of structures of
  the species $R(N)$ knowing that we already have a representation for
  $R$ and $N$. What we can do is to literally substitute arbitrarily
  selected $N$-structures in place of each vertex of an $R$-structure,
  For this, one can imagine that the vertices of the $R$-structure
  blow up into cells containing the $N$-structures.  The underlying
  set of a cellular configuration is the sum of the underlying sets of
  the structures contained in the cells. Thus, if we substitute the
  species of circular permutations into the species of trees, we
  obtain a species whose general type can be represented as in Figure
  4.

This representation is not the only one, and it is convenient to adapt to
the particularities of a species. For example, suppose that the species $N$
is \emph{pointed}, that is, equipped with a morphism $N \to^p B$ where $B$ is the species
of ``vertices'' ($B [E] = E$ for $E \in B$). Each structure $s \in N[E]$ then has
a \emph{base point} $p (s) \in E$. One can used the base points to
give another representation of $R (N)$-structures: for each vertex
of an $R$-structure one chooses an $N$-structure and makes the vertex \emph{coincide}
with the base point of the selected $N$-structure. For example, the base point
a rooted tree is the root; if we substitute the species $A$ of rooted trees
into that of (nonempty) linear orders, we obtain the species of vertebrates
(Ex. 9). When $N$ is pointed we can define substitution as follows: an
$R (N)$-structure on $E$ is a triplet $(v, \alpha, \beta)$, where
\begin{enumerate}
\item $v$ is an idempotent function $E \to E$,
\item $\alpha$ is a function that selects for each $x \in \Im (v)$ a
  structure $\alpha (x) \in N [v^{-1}\{x\}]$ such that the base point
  of $\alpha (x)$ \emph{coincides} with $x \in v^{-1}\{x\}$,
\item $\beta$ is an $R$-structure on $\Im (v)$.
\end{enumerate}
\end{rem}

FIGURE 4


\begin{ex}
Let $D$ be the species of endofunctions, $S$ that of
permutations, and $A$ that of rooted trees. We have the decomposition
\[ D = S(A). \] Indeed, let $\phi \in D [E]$. A point $x \in E$ is
\term{periodic} if there exists an integer $n \geq 1$ such that
$\phi^n(x) = x$. The function $\phi$ \term{permutes} the periodic
points.  For each $x \in E$ let $v (x)$ be the first periodic point in
the sequence $x, \phi(x), \phi^2(x) \dots$. The function $v$ is
idempotent and its image is the set of periodic points. For each $x
\in \Im (v)$ the fiber $v^{-1}\{x\}$ is equipped with a rooted tree
structure whose root is $x$. \emph{Conversely}, if one has an assembly
of rooted trees and a permutation of the set of roots, it is clear
that one can construct a corresponding endofunction $\phi$ (FIG 5)
\end{ex}






FIGURE 5

Examples 9 and 12 give a simple proof of Cayley's theorem: the number
$a_n$ of trees on a set of cardinality $n \geq 1$ is
$n^{n-2}$. Indeed, the number of vertebrates (bipointed trees) is
equal to $n^2a_n$.  Example 9 shows that the vertebrates are
\emph{linear} assembies of rooted trees.  Example 10 shows that
endofunctions are \emph{permuted} assemblies of rooted trees.  As the
number of linear orders coincides with the number of permutations, one
obtains $n^2 a_n = n^n$ $(n \geq 1)$.

Theorem 1 suggests adopting the notation $M (X)$ to designate a
species $M$. The variable $X$ is interpreted as the \emph{singleton}
species: there is only a single structure of the species $X$ (up to
isomorphism) and it is supported by the singletons. (???) The result
$M (X)$ of the substitution of $X$ in $M$ ​​is isomorphic to $M$. For
some species we adopt a frankly (???)  algebraic notation if it does
not create ambiguity. Thus, $\exp (X)$ or $e^X$ designate the uniform
species, $X e^X$ the species of pointed sets, $e^X - 1$ the uniform
nonempty species, $\cosh (X)$ and $\sinh (X)$ the uniform even and odd
species, $1/(1 - X)$ the species of linear orders, \etc However, we
retain the notation $S (X)$ to designate the species of permutations
in order to avoid confusion with $1/(1-X)$. Similarly, we will use the
notation $C (X)$ rather than $\log 1/(1-X)$ to designate the species
of circular permutations, \etc

\begin{ex}
  A preorder relation $\leq$ on $E$ determines an equivalence
  relation: $x \equiv y$ if and only if one has $x \leq y$ and $y \leq
  x$.  The preorder relation induced on the quotient $E / \equiv$ is
  an order relation (???), and conversely.  This shows that the
  species of preorders is obtained by substituting the species $e^X -
  1$ in the species of order relations. In particular, the species of
  total preorders is
  \[ \frac{1}{1 - (e^X - 1)} = \frac{1}{2 - e^X}. \]
\end{ex}

FIGURE 6

\begin{ex}
  In a graph, two vertices are \term{doubly connected} if we can
  connect them with a path avoiding any edge selected beforehand. The
  vertices of a graph can be partitioned into doubly connected
  components.  Let $\Gc$ be the species of connected pointed graphs
  and $\Gcc$ that of doubly connected pointed graphs. We have the
  relation
  \[ \Gc = \Gcc(Xe^{\Gc(X)}). \] Indeed, in a connected pointed graph,
  consider the doubly connected component $H$ of the base point; for
  each vertex $x$ let $v(x)$ be the closest vertex located in the
  component $H$. We can easily verify that $v$ is well defined. It is
  an idempotent function whose fibers are equipped with a structure of
  the species $X \cdot e^{\Gc(X)}$; the image of $v$ coincides with
  the doubly connected pointed graph $H$ (XXX FIG 6).
\end{ex}


FIGURE 7



FIGURE 8

\begin{ex}[Polya XXX CITE]
  Consider the species $A$ of rooted trees.  One has the identity (XXX
  FIG. 7)
\[ A = X \cdot \exp (A). \]
More generally, the species $A_R$ of R-enriched trees satisfies
the equation (see 1.3, Fig. 0 XXX)
\[ A_R = X \cdot R(A_R). \]
\end{ex}

\subsection{Differential calculus}

For any finite set $E$, let $E^+$ be the set obtained by adjoining to $E$
an additional item $*$:
\[ E^+ = E + \{ * \}. \]

\begin{defn}
The \term{derivative} species $M'$ of a species $M$ is defined as
follows:
\[ M' [E] = M [E^+]. \]
\end{defn}

\begin{ex}
  The derivative of the species $C (X)$ of circular permutations is
  the species of linear orders (Fig. 8 XXX):
  \[ C'(X) = \frac{1}{1 - X} \]
\end{ex}

FIGURE 9





FIGURE 10

\begin{ex}
  The derivative of the species of trees is that of forests of pointed
  trees XXX FIg 9).
\end{ex}

\begin{ex}
  Recall that a graph is \term{even} if the number of edges adjacent
  to each vertex is even. The derivative of the species of even graphs
  is the species of graphs (Fig. l0 XXX). (Harary and Palmer [15] XXX
  CITE)
\end{ex}

\begin{ex}
  The derivative of the species of linear orders $1 / (1-X)$ is equal
  to $(1 / (1-X)) \cdot (1 / (1-X))$.
\end{ex}

Recall that a \term{pointed} structure of the species $M$ on $E$ is an
element of $E \times M [E]$. We denote by $\pointed M$ the species of
pointed $M$-structures.

\bibliographystyle{plainnat}
\bibliography{series-formelles}

\end{document}
